---
title: "Oh_ProjectII"
author: "A.Oh"
date: '2022-04-05'
output: html_document
---
Purpose of my Project Part II:
The purpose of my project part II is to draw statistical analysis from the summary statistics I have extracted in Part 1 through comparisons of means t-test, chi squared test of proportions, and linear regression, as well as answer the question: what makes an academically successful student/individual?

Calling all dataframes created in part I of the project:
```{r}
#also installing packages I need for part II of the project
library("readxl")
library(tidyverse)
library(ggplot2)
library(magrittr)
library(dplyr)

#dataframe outlining individuals and their corresponding socioeconomic statuses
socecon <- read.csv('Data/socecon.csv')
head(socecon)

#dataframe outlining family sizes
famsize <- read.csv('Data/famsize.csv')
head(famsize)

#dataframe outlining the socioeconomic status, family size, and urbanicity
sesfams <- read.csv('Data/sesfams.csv')
head(sesfams)

#dataframe outlining the time students work on homework outside of school on average weekly
hwtime <- read.csv('Data/hwtime.csv')
head(hwtime)

#dataframe below outline the achievements in each subjects for each student
reading <- read.csv('Data/table_reading.csv')
math <- read.csv('Data/table_math.csv')
science <- read.csv('Data/table_science.csv')
socials <- read.csv('Data/table_ss.csv')
head(reading)
head(math)
head(science)
head(socials)

#dataframe outlines the grades of students in all subjects (all in one dataframe)
grades <- read.csv('Data/table_grades.csv')
head(grades)

#dataframe outlines the grades of students and the time students take on homework outside of school
grades_time <- read.csv('Data/grades_time.csv')
head(grades_time)

#dataframe outlining the AP courses offered for each region in America
apf <- read.csv('Data/apf.csv')
head(apf)

#dataframe outlining the marital status of the individual's parents
marital <- read.csv('Data/marital.csv')
head(marital)

#dataframe outlining if students had taken advanced math in eighth grade and are smoking cigarettes right now
adsm <- read.csv('Data/marital.csv')
head(marital)

#dataframe outlining the expected income of students
expinc <- read.csv('Data/expinc.csv')
head(expinc)

#dataframe outlining the time spent on extracurricular activities and the highest level of education students expect to achieve
eced <- read.csv('Data/eced.csv')
head(eced)

#dataframe outlining the different types of schools attended by individuals in eighth grade and high school
table_school <- read.csv('Data/table_school.csv')
head(table_school)

#dataframe outlining the frequency individuals who took an advanced program after taking advanced math in eighth grade
apadm <- read.csv('Data/apadm.csv')
head(apadm)

```

Calling the original nels dataset to make 2x2 count tables for each test:
```{r}
nels <- read_excel('Data/nels.xlsx')
head(nels)
```

Below, I will be conducting a comparisons of means t-test for four different variables (two pairs):
```{r}
#first part variables will be trying to see if there is an association between the socioeconomic status and if the student takes AP classes in high school

#creating a separate dataframe to get rid of inappropriate data variables from columns
socap <- nels%>%
            select(`Case Number`, `Socio-Economic Status`,`Advanced Placement Program Taken?`) %>%
            filter(`Advanced Placement Program Taken?` != "8.0")
print(socap)

str(socap)

socap_1 <- table(socap$`Socio-Economic Status`,socap$`Advanced Placement Program Taken?`)
addmargins(socap_1)

#creating boxplot to compare spread
boxplot(`Socio-Economic Status`~`Advanced Placement Program Taken?`, data=socap, main='Spread of Socioeconomic status by Advanced Placement Program Taken', xlab='Socioeconomic Status', col='purple2', ylab='AP Program Taken?',border = 'purple3', horizontal=TRUE)

#conducting t-test
t.test(`Socio-Economic Status`~`Advanced Placement Program Taken?`, data=socap, equal.var=FALSE)

#second part variables will be trying to see if there is an association between expected income at 30 and whether or not the student has binged on alcohol ever

#(summary plot of binging on alcohol)
alc <- nels %>%
        select(`Case Number`,`Binged on Alcohol Ever?`)
print(alc)

#(plot of binging on alcohol)
ggplot(alc,aes(`Binged on Alcohol Ever?`)) + geom_bar(colour = "white", fill = "pink") + ggtitle("Whether or not Student has Binged on Alcohol")

#percentage dataframe for alcohol
summary_alc <- alc %>%
  group_by(alc$`Binged on Alcohol Ever?`) %>%
  summarise(cnt = n()) %>%
  mutate(freq = formattable::percent(cnt / sum(cnt))) %>% 
  arrange(desc(freq))

print(summary_alc)

#printing the dataframe expected to reference
print(expinc)

expinc_1 <- table(expinc$no_outlier_expected_income_30,expinc$new_alc)
addmargins(expinc_1)

#creating boxplot to compare spread
boxplot(no_outlier_expected_income_30~new_alc, data=expinc, main='Spread of Expected Income by age 30 by Alcohol Binging', xlab='Expected Income', col='blue2', ylab='Binging on Alcohol?',border = 'blue3', horizontal=TRUE)

#conducting t-test
t.test(no_outlier_expected_income_30~new_alc, data=expinc, equal.var=FALSE)

```


Below, I will be conducting a chi-squared test for proportions for six variables (three pairs):
```{r}
#first part variables are trying to see if there is an association between the education achievements in students who took AP classes and taking advanced math in eighth grade

#the 2x2 subset table
admap <- table(apadm$Advanced.Placement.Program.Taken.,apadm$Advanced.Math.Taken.in.Eighth.Grade)
print(admap)
addmargins(admap)

#displaying the dataframe for this test
print(apadm)

#performing the chi-squared test
chisq.test(apadm$Advanced.Placement.Program.Taken.,apadm$Advanced.Math.Taken.in.Eighth.Grade)

#second part variables are trying to see if there is an association between weekly time spent on homework outside of school and expected highest level of education achieved

#the 2x2 table for hwtime and highest level of education achieved
hwtime_ed <- table(nels$`Time Spent on Homework out of School per Week in Twelfth Grade`,nels$`Highest level of education expected`)
addmargins(hwtime_ed)

#performing the chi-squared test
chisq.test(nels$`Time Spent on Homework out of School per Week in Twelfth Grade`,nels$`Highest level of education expected`)

#third part variables are trying to see if there is an association between highest level of education expected to achieve and urbanicity of students

#the 2x2 table for highest level of education and urbanicity of students
ed_urban <- table(nels$`Highest level of education expected`,nels$Urbanicity)
addmargins(ed_urban)

#performing the chi-squared test
chisq.test(nels$`Highest level of education expected`,nels$Urbanicity)

```
Below, I will be conducting a linear regression for two variables:
```{r}
#first part variables will be trying to see if there is an association between socioeconomic status and expected income at age 30
socecon_inc <- data.frame(expinc$no_outlier_case_num,expinc$no_outlier_expected_income_30,expinc$new_socecon)
head(socecon_inc)

socecon_exp <- data.frame(socecon,nels$`Expected income at age 30`)
head(socecon_exp)

#create a scatterplot for expected income with socioeconomic status
ggplot(data=socecon_exp) + geom_point(mapping = aes(x = Socio.Economic.Status , y = nels..Expected.income.at.age.30.)) + ggtitle("Socioeconomic status and Expected Income at 30") + xlab("Socioeconomic status")+ ylab("Expected Income at 30")

#(without outliers)
ggplot(data=socecon_inc) + geom_point(mapping = aes(x = expinc.new_socecon , y = expinc.no_outlier_expected_income_30)) + ggtitle("Socioeconomic status and Expected Income at 30") + xlab("Socioeconomic status")+ ylab("Expected Income at 30")

#correlation between expected income and socioeconomic status
corr_socinc <- cor(socecon_inc$expinc.new_socecon,socecon_inc$expinc.no_outlier_expected_income_30)
print(corr_socinc)

r_sq <- corr_socinc^2
print(r_sq)

#plot the linear regression line on the scatterplot
ggplot(data=socecon_inc, aes(x = expinc.new_socecon , y = expinc.no_outlier_expected_income_30)) + geom_point() + geom_smooth(method="lm") + ggtitle("Socioeconomic status and Expected Income at 30") + xlab("Socioeconomic status")+ ylab("Expected Income at 30")

#finding lm model
socecon_inc_model <- lm(expinc.new_socecon~expinc.no_outlier_expected_income_30, data = socecon_inc)
print(socecon_inc_model)

#finding the residuals and plotting them
residuals <- resid(socecon_inc_model)
print(residuals)

hist(residuals) #(histogram of the residuals)

qqnorm(residuals) #plotting the normal probability plot of the residuals

plot(fitted(socecon_inc_model),residuals)#creating a residual plot

std_res <- rstandard(socecon_inc_model)
print(std_res) #calculating standards residuals

plot(fitted(socecon_inc_model),std_res) #plot of standardized residual plot

#Combining regression with standard residuals
socecon_expinc <- socecon_inc[c("expinc.new_socecon", "expinc.no_outlier_expected_income_30")]
soc_inc_std <- cbind(socecon_expinc, std_res)
print(soc_inc_std)

#Removing standard residuals that are greater than 2
socecon_inc_analysis <- soc_inc_std[soc_inc_std$std_res < 2,]
print(socecon_inc_analysis)

#results of the linear regression
summary(lm(expinc.new_socecon~expinc.no_outlier_expected_income_30, data = socecon_inc))

```

